Top Mortality causes in young adults and middle age adults in Switzerland

### 2024-04-21 paper/paper.qmd
Update typos and references

---
title: "Top Mortality causes in young adults and middle age adults in Switzerland"
subtitle: "Analyzing Switzerland Mortality causes data in 2010, 2015, adn 2019 from WHO database"
author: 
  - Thu Dong
thanks: "Code and data are available at: https://github.com/thu2912/Top-mortality-causes-of-age-groups-in-Switzerland.git"
date: today
date-format: long
abstract: "Motivated by the lack of findings on different mortality causes for different age groups, this paper examines the main causes of mortality among young adults and middle-aged adults in Switzerland using data from the WHO Mortality Database for the years 2010, 2015, and 2019. The paper found that the leading causes of death for young adults are self-harm and drug use disorders, while middle-aged adults are most affected by ischemic heart disease and trachea, bronchus, and lung cancers. Moreover, it is also found that one of the main mortality causes in both age groups is self-harm which highlights the importance of mental health."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

## Clean Data
he data utilized for this paper is downloaded manually through the WHO Mortality Database. All the data analysis was done through R programming language (@citeR) with the aid of the following packages:rstanarm (@rstanarm), tidyverse (@tidyverse), knitr (@knitr), modelsummary (@modelsummary), ggplot2 (@ggplot2), dplyr (@dplyr), parameters (@parameters), broom (@broom), broom.mixed (@broommixed).

@@

## Leading cause of death in middle-aged people in Switzerland. 
As shown in @tbl-tbl4, Ischaemic heart disease (IHD) is the leading cause of death in middle-aged adults. Notably, the leading cause of death for the whole population in Switzerland is Ischaemic heart disease with approximately 11,000 cases per year since 2010. (“Coronary Heart Disease in Switzerland” 2020). This disease is not only common in Switzerland, but it is also common worldwide, It results from reduced blood flow to the heart muscle due to narrowed or blocked coronary arteries. Some risk factors for this disease include high blood pressure, high cholesterol, smoking, diabetes, and obesity. It is reported that IHD is responsible for approximately 16% of total deaths worldwide1.(World 2020).

One of the reasons why IHD is the highest cause of death in Switzerland is because of the citizen’s tobacco usage. It is known that tobacco usage is one of the risk factors for IHD. In 2017, Swiss smokers used an average of 9.5 cigarettes per day. Additionally, approximately 21% of smokers consumed at least 20 cigarettes per day. Overall, 6% of the Swiss population aged 15 and over were smokers. (“Level of Use - Swiss Association for Tobacco Control” 2016). It is shown that smokers are 2 or 4 times more likely to develop IHD than non-smokers. With these statistics in mind, policymakers passed regulations such as banning smoking indoors and banning selling tobacco to people under 18.

@@

Switzerland witnessed an alarming rate of self-harm incidents. A government-commissioned study in the French-speaking part of Switzerland documented 554 cases of self-harm by 490 patients over a ten-month period. Additionally, half of the patients were aged 18 to 34, and 65% of them were reported to have a difficult socioeconomic situation. (SWI swissinfo.ch 2019). In 2016, Switzerland launched a national action plan on self- harm/ suicide prevention with some key measures: raise awareness about these issues, provide accessible help, recognize early behavior of self-harm and timely intervene, reduce means of suicide and self-harm, etc.

@@

In the future, we will extend the study to include more years and possibly conduct a time-series analysis to observe trends over a continuous period. This change can also help us study for effect of different policies on mortality rates. Additionally, we can also compare the mortality causes and trends in Switzerland with those in other countries with similar socioeconomic conditions. As a result, we can identify the unique factors that make Switzerland one of the countries with the longest life expectancy.

>>>>

---
title: "Top Mortality causes in young adults and middle age adults in Switzerland"
subtitle: "Analyzing Switzerland Mortality causes data in 2010, 2015, and 2019 from WHO database"
author: 
  - Thu Dong
thanks: "Code and data are available at: https://github.com/thu2912/Top-mortality-causes-of-age-groups-in-Switzerland.git"
date: today
date-format: long
abstract: "Motivated by the lack of findings on different mortality causes for different age groups, this paper examines the main causes of mortality among young adults and middle-aged adults in Switzerland using data from the WHO Mortality Database for the years 2010, 2015, and 2019. The paper found that the leading causes of death for young adults are self-harm and drug use disorders, while middle-aged adults are most affected by ischemic heart disease and trachea, bronchus, and lung cancers. Moreover, it is also found that one of the main mortality causes in both age groups is self-harm which highlights the importance of mental health."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

## Clean Data
The data utilized for this paper is downloaded manually through the WHO Mortality Database. All the data analysis was done through R programming language (@citeR) with the aid of the following packages:rstanarm (@rstanarm), tidyverse (@tidyverse), knitr (@knitr), modelsummary (@modelsummary), ggplot2 (@ggplot2), dplyr (@dplyr), parameters (@parameters), broom (@broom), broom.mixed (@broommixed).

@@

## Leading cause of death in middle-aged people in Switzerland. 
As shown in @tbl-tbl4, Ischaemic heart disease (IHD) is the leading cause of death in middle-aged adults. Notably, the leading cause of death for the whole population in Switzerland is Ischaemic heart disease with approximately 11,000 cases per year since 2010. (@worldlife2020coronary). This disease is not only common in Switzerland, but it is also common worldwide, It results from reduced blood flow to the heart muscle due to narrowed or blocked coronary arteries. Some risk factors for this disease include high blood pressure, high cholesterol, smoking, diabetes, and obesity. It is reported that IHD is responsible for approximately 16% of total deaths worldwide.(World 2020).

One of the reasons why IHD is the highest cause of death in Switzerland is because of the citizen’s tobacco usage. It is known that tobacco usage is one of the risk factors for IHD. In 2017, Swiss smokers used an average of 9.5 cigarettes per day. Additionally, approximately 21% of smokers consumed at least 20 cigarettes per day. Overall, 6% of the Swiss population aged 15 and over were smokers.(@swiss2016tobacco) It is shown that smokers are 2 or 4 times more likely to develop IHD than non-smokers. With these statistics in mind, policymakers passed regulations such as banning smoking indoors and banning selling tobacco to people under 18.

@@

Switzerland witnessed an alarming rate of self-harm incidents. A government-commissioned study in the French-speaking part of Switzerland documented 554 cases of self-harm by 490 patients over a ten-month period. Additionally, half of the patients were aged 18 to 34, and 65% of them were reported to have a difficult socioeconomic situation. (@swissinfo2019selfharm). In 2016, Switzerland launched a national action plan on self- harm/ suicide prevention with some key measures: raise awareness about these issues, provide accessible help, recognize early behavior of self-harm and timely intervene, reduce means of suicide and self-harm, etc.

@@

In the future, we will extend the study to include more years and possibly conduct a time-series analysis to observe trends over a continuous period. This change can also help us study for effect of different policies on mortality rates. Additionally, we can also compare the mortality causes and trends in Switzerland with those in other countries with similar socioeconomic conditions. As a result, we can identify the unique factors that make Switzerland one of the countries with the highest life expectancy.
